Negative Types

### 2014-05-23 ring.tex
more about paths and groupoids

\newcommand{\bdim}[1]{\textsf{dim}(#1)}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\proves}{\vdash}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
\newcommand{\adjoint}[1]{#1^{\dagger}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\plus}{\raisebox{.4\height}{\scalebox{.6}{+}}}
\newcommand{\minus}{\raisebox{.4\height}{\scalebox{.8}{-}}}
\newcommand{\mm}{\texttt{-}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\negp}[1]{\textit{neg}(#1)}
\newcommand{\inl}[1]{\textsf{inl}~#1}
\newcommand{\inr}[1]{\textsf{inr}~#1}
\newcommand{\lolli}{\multimap} 
\newcommand{\cubt}{\mathbb{T}}
\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\nodet}[2]{\fcolorbox{black}{white}{$#1$}\fcolorbox{black}{gray!20}{$#2$}}
\newcommand{\hast}{:\mkern -2.5mu:\;}

@@

%% double-blind reviewing...
\title{Polarized Cubical Types}
\authorinfo{}{}{}
\maketitle

@@

The \textbf{Int} construction or the $\mathcal{G}$ construction are neat. As
Neel K. explains, given first-order types and feedback you get higher-order
functions. But if you do the construction on the additive structure, you lose
the multiplicative structure. It turns out that this is related to a deep
open problem in algebraic topology and homotopy theory that was recently
solved. We ``translate'' that solution to a computational type-theoretic
world. This has evident connections to homotopy (type?) theory that remain to
be investigated.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Problem and the Intuitive Solution}

We lose the multiplicative structure. No evident way to define the
multiplication functor. Perhaps there is a clever way. But this turns out to
be a well-known open problem. Review the problem and tell the story from the
algebraic topology perspective. Regular first-order types are viewed as
$0$-dimensional cubes. Generalize to $n$-dimensional cubes.

@@

Our cubical types $\tau$ include the empty type 0, the unit type 1,
conventional sum and product types, as well as \emph{negative} types:
\[\begin{array}{rcl}
\tau &::=& 0 \alt 1 \alt \tau_1 + \tau_2 \alt \tau_1 * \tau_2 \alt - \tau
\end{array}\]
We use $\tau_1 - \tau_2$ to abbreviate $\tau_1 + (- \tau_2)$ and more
interestingly $\tau_1 \lolli \tau_2$ to abbreviate $(- \tau_1) + \tau_2$.
The \emph{dimension} of a type is defined as follows:
\[\begin{array}{rcl}
\bdim{\cdot} &\hast& \tau \rightarrow \mathbb{N} \\
\bdim{0} &=& 0 \\
\bdim{1} &=& 0 \\
\bdim{\tau_1 + \tau_2} &=& \max(\bdim{\tau_1},\bdim{\tau_2}) \\
\bdim{\tau_1 * \tau_2} &=& \bdim{\tau_1} + \bdim{\tau_2} \\
\bdim{- \tau} &=& \max(1,\bdim{\tau})
\end{array}\]
The base types have dimension 0. Using negative types but no product types
raises the dimension to 1. This is the situation with the \textbf{Int} or
$\mathcal{G}$ construction. Once product types are used, the dimension can
increase without bounds.

\begin{figure*}
\[\begin{array}{c}
\nodet{S_1}{S_2}
\quad\otimes\quad
\nodet{(\nodet{S_3}{S_4})}{(\nodet{S_5}{S_6})} \quad= \\
\\
\nodet{(\nodet{{(\nodet{S_1 \times S_3}{S_1 \times S_4})}}
              {{(\nodet{S_1 \times S_5}{S_1 \times S_6})}})}
      {(\nodet{{(\nodet{S_2 \times S_3}{S_2 \times S_4})}}
              {{(\nodet{S_2 \times S_5}{S_2 \times S_6})}})}
\end{array}\]
\\
\\
\begin{tikzpicture}
\node[above] at (0,0) {\pp};
\node[left] at (0,0) {$S_1$};
\draw[fill] (0,0) circle [radius=0.05];
\node[above] at (0.6,0) {\mm};
\node[right] at (0.6,0) {$S_2$};
\draw[fill] (0.6,0) circle [radius=0.05];
\draw[-] (0,0) -- (0.6,0);
\node at (1.6,0) {$\otimes$};

%%
\node[left] at (2.5,-0.5) {\pp\pp};
\node[below] at (2.5,-0.5) {$S_3$};
\draw[fill] (2.5,-0.5) circle [radius=0.05];
\node[right] at (3.5,-0.5) {\pp\mm};
\node[below] at (3.5,-0.5) {$S_4$};
\draw[fill] (3.5,-0.5) circle [radius=0.05];
\draw[-] (2.5,-0.5) -- (3.5,-0.5);
\draw[-] (2.5,-0.5) -- (2.5,0.5);
\node[left] at (2.5,0.5) {\mm\pp};
\node[above] at (2.5,0.5) {$S_5$};
\draw[fill] (2.5,0.5) circle [radius=0.05];
\node[right] at (3.5,0.5) {\mm\mm};
\node[above] at (3.5,0.5) {$S_6$};
\draw[fill] (3.5,0.5) circle [radius=0.05];
\draw[-] (2.5,0.5) -- (3.5,0.5);
\draw[-] (3.5,-0.5) -- (3.5,0.5);
%% 
\node at (5,0) {$=$};
%% 
\node[left] at (7.5,0.75) {$(S_2 \times S_3)$\,\mm\pp\pp};
\draw[fill] (7.5,0.75) circle [radius=0.05];
\node[right] at (9.5,0.75) {$(S_2 \times S_4)$\,\mm\pp\mm};
\draw[fill] (9.5,0.75) circle [radius=0.05];
\node[above right] at (10.2,1.2) {$(S_2 \times S_6)$\,\mm\mm\mm};
\draw[fill] (10.2,1.2) circle [radius=0.05];
\node[above left] at (8.2,1.2) {$(S_2 \times S_5)$\,\mm\mm\pp};
\draw[fill] (8.2,1.2) circle [radius=0.05];
%%
\node[left] at (7.5,-0.75) {$(S_1 \times S_3)$\,\pp\pp\pp};
\draw[fill] (7.5,-0.75) circle [radius=0.05];
\node[right] at (9.5,-0.75) {$(S_1 \times S_4)$\,\pp\pp\mm};
\draw[fill] (9.5,-0.75) circle [radius=0.05];
\node[above right] at (10.2,-0.3) {$(S_1 \times S_6)$\,\pp\mm\mm};
\draw[fill] (10.2,-0.3) circle [radius=0.05];
\node[left] at (8.2,-0.3) {$(S_1 \times S_5)$\,\pp\mm\pp};
\draw[fill] (8.2,-0.3) circle [radius=0.05];
%%
\draw[-] (7.5,0.75) -- (9.5,0.75);
\draw[-] (9.5,0.75) -- (10.2,1.2);
\draw[-] (10.2,1.2) -- (8.2,1.2);
\draw[-] (8.2,1.2) -- (7.5,0.75);
%%
\draw[-] (7.5,-0.75) -- (9.5,-0.75);
\draw[-] (9.5,-0.75) -- (10.2,-0.3);
\draw[-,dashed] (10.2,-0.3) -- (8.2,-0.3);
\draw[-,dashed] (8.2,-0.3) -- (7.5,-0.75);
%%
\draw[-] (7.5,0.75) -- (7.5,-0.75);
\draw[-] (9.5,0.75) -- (9.5,-0.75);
\draw[-] (10.2,1.2) -- (10.2,-0.3);
\draw[-,dashed] (8.2,1.2) -- (8.2,-0.3);
\end{tikzpicture}
\caption{\label{mult}Example of multiplication of two cubical types.}
\end{figure*}

This point is made precise in the denotation of types which maps a type of
dimension $n$ to $n$-dimensional cube. We represent such a cube syntactically
as a binary tree of maximum depth~$n$ with nodes of the form
$\nodet{\cubt_1}{\cubt_2}$. In such a node, $\cubt_1$ is the positive
subspace and $\cubt_2$ (shaded in gray) is the negative subspace along the
first dimension. Each of these subspaces is itself a cube of a lower
dimension. The $0$-dimensional cubes are plain sets representing the
denotation of conventional first-order types. We use $S$ to denote the
denotations of these plain types. A 1-dimensional cube, $\nodet{S_1}{S_2}$,
intuitively corresponds to the difference $t_1 - t_2$ of the two types whose
denotations are $S_1$ and $S_2$ respectively. The type can be visualized as a
``line'' with polarized endpoints connecting the two points $S_1$ and
$S_2$. A full 2-dimensional cube,
$\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$, intuitively corresponds to
the iterated difference of the appropriate types $(t_1-t_2)-(t_3-t_4)$ where
the product of the successive signs from the outermost box encodes the
sign. The type can be visualized as a ``square'' with polarized corners
connecting the two lines corresponding to $(t_1-t_2)$ and $(t_3-t_4)$. (See
Fig.~\ref{mult} which is further explained after we discuss multiplication
below.)

@@

%%%%%%%%%%%%%%%%%%%%
\subsection{Type Isomorphisms}

Our (simple) semantics of types identifies several structurally different
types such as $(1+(1+1))$ and $((1+1)+1)$. In some sense, this is innocent as
the types are isomorphic. However, in the operational semantics discussed in
the next section, we make the computational content of such type isomorphisms
explicit. Some other isomorphic types like $(t_1*t_2)$ and $(t_2*t_1)$ map to
different cubes and are \emph{not} identified: explicit isomorphisms are
needed to mediate between them. In most cases, the isomorphisms are easy to
define except for the isomorphisms involving the empty type. In particular,
if negative types are to be interpreted as their name suggests, we must have
an isomorphism between $(t-t)$ and the empty type 0. Semantically the former
denotes $\nodet{\cubt}{\cubt}$ and the latter denotes the empty set. Their
denotations are different and there is no way so far to express the fact that
these two cubes should be identified. Intuitively, what is needed is the
ability to contract a line whose endpoints are identical to a single
point. It should then follow that a square where one edge contracts to a
point would itself contract to a line, and a 3-dimensional cube in which one
face contracts to a point would itself contract to a square, etc. This is, of
course, where the ideas of homotopy (type) theory enter the development.

Consider the situation with the simplest instance of the desired isomorphism
which is between $(1-1)$ and the empty type:
\begin{center}
\begin{tikzpicture}
\node[above] at (0,0) {\pp};
\node[left] at (-0.2,0) {$1$};
\draw[fill] (0,0) circle [radius=0.05];
\node[above] at (2.6,0) {\mm};
\node[right] at (2.8,0) {$1$};
\draw[fill] (2.6,0) circle [radius=0.05];
\draw[->,red] (0.1,0) -- (2.5,0);
\draw[->,red] (-0.1,-0.1) arc (135:405:0.2);
\node[below,red] at (0,-0.4) {\textsf{refl}};
\draw[->,red] (2.5,-0.1) arc (135:405:0.2);
\node[below,red] at (2.6,-0.4) {\textsf{refl}};
\node[above] at (1.3,0) {$I$};
\node[right] at (1.3,-0.5) {$q$};
%%
\node[below] at (1.3,-1.5) {$\emptyset$};
\draw[fill] (1.3,-1.5) circle [radius=0.05];
\draw[->,red] (1.4,-1.4) arc (-45:225:0.2);
\node[right,red] at (1.5,-1.3) {\textsf{refl}};
%% 
\draw[->,double,blue] (1.3,-0.1) -- (1.3,-1);
\end{tikzpicture}
\end{center}
The top of the figure is the 1-dimensional cube representing the type $(1-1)$
except that we view the edge connecting the two endpoints as a \emph{path}
$I$ that is initially non-contractible, i.e., it is different from the
identity paths labeled \textsf{refl} that are the trivial paths asserting
that $1$ is contractible to itself. The bottom of the figure is the
0-dimensional cube representing the empty type. There is a trivial path
\textsf{refl} connecting that point to itself. To express the equivalence of
$(1-1)$ to 0, we add a 2-path, i.e. a path between paths, that asserts the
equivalence of the path $I$ with \textsf{refl} on the empty type. That
effectively contracts the two points and collapses them to a single
point. Surprisingly, that is everthing that we need. The extension to higher
dimensions just ``works'' because paths in homotopy type theory have a rich
structure. We explain the details below.

\begin{verbatim}
So the story is that as we are building
the denotations of types, we add these
paths making up the cubes. The only
case where we add new paths is multiplication
(see Fig.1 again). We take the paths p in S1
and the paths q in S2 and build paths (p,q)
in the product space.

@@

Now that the type structure is defined, we turn our attention to the notion
of values. Intuitively, a value of the $n$-dimensional type $\tau$ is an
element of one of the sets located in one of the corners of the
$n$-dimensional cube denoted by $\tau$. Thus to specify a value, we must
first specify one of the corners of the cube (or equivalently one of the
leaves in the binary tree representation) which can easily be done using a
sequence $p$ of $+$ and $-$ polarities indicating how to navigate the cube in
each successive dimension starting from a fixed origin to reach the desired
corner. We write $v^{p}$ for the value $v$ located at corner $p$ of the cube
associated with its type. We use $\epsilon$ for the empty sequence of
polarities and identify $v$ with $v^\epsilon$. Note that the polarities
doesn't completely specify the type since different types like $(1+(1+1))$
and $((1+1)+1)$ are assigned the same denotation. What the path $p$ specifies
is the \emph{polarity} of the value, or its ``orientation'' in the space
denoted by its type. Formally:
\[\begin{array}{c}
\infer{() : 1}{} \\
\qquad
\infer[\textit{neg}]{v^{\negp{p}} : - \tau}{v^p : \tau} 
\\
\infer[\textit{left}]{(\inl{v})^{p} : \tau_1 + \tau_2}{v^{p} : \tau_1}
\qquad
\infer[\textit{right}]{(\inr{v})^{p} : \tau_1 + \tau_2}{v^{p} : \tau_2} 
\\
\infer[\textit{prod}]{(v_1,v_2)^{p_1 \cdot p_2} : \tau_1 * \tau_2}
      {v_1^{p_1} : \tau_1 & v_2^{p_2} : \tau_2} 
\end{array}\]
The rules \textit{left} and \textit{right} reflect the fact that sums do not
increase the dimension. Note that when $p$ is $\epsilon$, we get the
conventional values for the 0-dimensional sum type. The rule \textit{prod} is
the most involved one: it increases the dimension by \emph{concatenating} the
two dimensions of its arguments. For example, if we pair $v_1^{\pp}$ and
$v_2^{\mm\pp}$ we get $(v_1,v_2)^{\pp\mm\pp}$. (See Fig.~\ref{mult} for the
illustration.) Note again that if both components are 0-dimensional, the pair
remains 0-dimensional and we recover the usual rule for typing values of
product types. The rule \textit{neg} uses the function below which states
that the negation of a value $v$ is the same value $v$ located at the
``opposite'' corner of the cube:
\[\begin{array}{rcl}
\negp{\epsilon} &=& \epsilon \\
\negp{+p} &=& -\negp{p} \\
\negp{-p} &=& +\negp{p}
\end{array}\]

@@

>>>>

\newcommand{\refl}[1]{\textsf{refl}_{#1}}
\newcommand{\seg}[1]{\textsf{seg}_{#1}}
\newcommand{\bdim}[1]{\textsf{dim}(#1)}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\proves}{\vdash}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
\newcommand{\adjoint}[1]{#1^{\dagger}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\plus}{\raisebox{.4\height}{\scalebox{.6}{+}}}
\newcommand{\minus}{\raisebox{.4\height}{\scalebox{.8}{-}}}
\newcommand{\mm}{\texttt{-}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\negp}[1]{\textit{neg}(#1)}
\newcommand{\inl}[1]{\textsf{inl}~#1}
\newcommand{\inr}[1]{\textsf{inr}~#1}
\newcommand{\lolli}{\multimap} 
\newcommand{\cubt}{\mathbb{T}}
\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\nodet}[2]{\fcolorbox{black}{white}{$#1$}\fcolorbox{black}{gray!20}{$#2$}}
\newcommand{\hast}{:\mkern -2.5mu:\;}

@@

%% double-blind reviewing...
\title{Negative Types}
\authorinfo{}{}{}
\maketitle

@@

The \textbf{Int} construction or the $\mathcal{G}$ construction are neat. As
Neel K. explains, given first-order types and feedback you get higher-order
functions. But if you do the construction on the additive structure, you lose
the multiplicative structure. It turns out that this is related to a deep
open problem in algebraic topology and homotopy theory that was recently
solved. We ``translate'' that solution to a computational type-theoretic
world. This has evident connections to homotopy (type) theory that remain to
be investigated.

@@

We lose the multiplicative structure. No evident way to define the
multiplication functor. Perhaps there is a clever way. But this turns out to
be a well-known open problem. Review the problem and tell the story from the
algebraic topology perspective. Regular first-order types are viewed as
$0$-dimensional cubes. The \textbf{Int} construction generalizes types to
1-dimensional cubes. Our work here generalizes types to $n$-dimensional
cubes.

@@

Our types $\tau$ include the empty type 0, the unit type 1, conventional sum
and product types, as well as \emph{negative} types:
\[\begin{array}{rcl}
\tau &::=& 0 \alt 1 \alt \tau_1 + \tau_2 \alt \tau_1 * \tau_2 \alt - \tau
\end{array}\]
We use $\tau_1 - \tau_2$ to abbreviate $\tau_1 + (- \tau_2)$ and more
interestingly $\tau_1 \lolli \tau_2$ to abbreviate $(- \tau_1) + \tau_2$.
The \emph{dimension} of a type is defined as follows:
\[\begin{array}{rcl}
\bdim{\cdot} &\hast& \tau \rightarrow \mathbb{N} \\
\bdim{0} &=& 0 \\
\bdim{1} &=& 0 \\
\bdim{\tau_1 + \tau_2} &=& \max(\bdim{\tau_1},\bdim{\tau_2}) \\
\bdim{\tau_1 * \tau_2} &=& \bdim{\tau_1} + \bdim{\tau_2} \\
\bdim{- \tau} &=& \max(1,\bdim{\tau})
\end{array}\]
The base types have dimension 0. If negative types are not used, all
dimensions remain at 0. If negative types are used but no product types of
negative types appear anywhere, the dimension is raised to 1. This is the
situation with the \textbf{Int} or $\mathcal{G}$ construction. Once negative
and product types are freely used, the dimension can increase without bounds.

\begin{figure*}
\[\begin{array}{c}
\nodet{S_1}{S_2}
\quad\otimes\quad
\nodet{(\nodet{S_3}{S_4})}{(\nodet{S_5}{S_6})} \quad= \\
\\
\nodet{(\nodet{{(\nodet{S_1 \times S_3}{S_1 \times S_4})}}
              {{(\nodet{S_1 \times S_5}{S_1 \times S_6})}})}
      {(\nodet{{(\nodet{S_2 \times S_3}{S_2 \times S_4})}}
              {{(\nodet{S_2 \times S_5}{S_2 \times S_6})}})}
\end{array}\]
\\
\\
\begin{center}
\begin{tikzpicture}
\node[above] at (0,0) {\pp};
\node[left] at (0,0) {$S_1$};
\draw[fill] (0,0) circle [radius=0.05];
\node[above] at (0.6,0) {\mm};
\node[right] at (0.6,0) {$S_2$};
\draw[fill] (0.6,0) circle [radius=0.05];
\draw[-,dotted] (0,0) -- (0.6,0);
\node at (1.6,0) {$\otimes$};

%%
\node[left] at (2.5,-0.5) {\pp\pp};
\node[below] at (2.5,-0.5) {$S_3$};
\draw[fill] (2.5,-0.5) circle [radius=0.05];
\node[right] at (3.5,-0.5) {\pp\mm};
\node[below] at (3.5,-0.5) {$S_4$};
\draw[fill] (3.5,-0.5) circle [radius=0.05];
\draw[-,dotted] (2.5,-0.5) -- (3.5,-0.5);
\draw[-,dotted] (2.5,-0.5) -- (2.5,0.5);
\node[left] at (2.5,0.5) {\mm\pp};
\node[above] at (2.5,0.5) {$S_5$};
\draw[fill] (2.5,0.5) circle [radius=0.05];
\node[right] at (3.5,0.5) {\mm\mm};
\node[above] at (3.5,0.5) {$S_6$};
\draw[fill] (3.5,0.5) circle [radius=0.05];
\draw[-,dotted] (2.5,0.5) -- (3.5,0.5);
\draw[-,dotted] (3.5,-0.5) -- (3.5,0.5);
%% 
\node at (5,0) {$=$};
%% 
\node[left] at (7.5,0.75) {$(S_2 \times S_3)$\,\mm\pp\pp};
\draw[fill] (7.5,0.75) circle [radius=0.05];
\node[right] at (9.5,0.75) {$(S_2 \times S_4)$\,\mm\pp\mm};
\draw[fill] (9.5,0.75) circle [radius=0.05];
\node[above right] at (10.2,1.2) {$(S_2 \times S_6)$\,\mm\mm\mm};
\draw[fill] (10.2,1.2) circle [radius=0.05];
\node[above left] at (8.2,1.2) {$(S_2 \times S_5)$\,\mm\mm\pp};
\draw[fill] (8.2,1.2) circle [radius=0.05];
%%
\node[left] at (7.5,-0.75) {$(S_1 \times S_3)$\,\pp\pp\pp};
\draw[fill] (7.5,-0.75) circle [radius=0.05];
\node[right] at (9.5,-0.75) {$(S_1 \times S_4)$\,\pp\pp\mm};
\draw[fill] (9.5,-0.75) circle [radius=0.05];
\node[above right] at (10.2,-0.3) {$(S_1 \times S_6)$\,\pp\mm\mm};
\draw[fill] (10.2,-0.3) circle [radius=0.05];
\node[left] at (8.2,-0.3) {$(S_1 \times S_5)$\,\pp\mm\pp};
\draw[fill] (8.2,-0.3) circle [radius=0.05];
%%
\draw[-,dotted] (7.5,0.75) -- (9.5,0.75);
\draw[-,dotted] (9.5,0.75) -- (10.2,1.2);
\draw[-,dotted] (10.2,1.2) -- (8.2,1.2);
\draw[-,dotted] (8.2,1.2) -- (7.5,0.75);
%%
\draw[-,dotted] (7.5,-0.75) -- (9.5,-0.75);
\draw[-,dotted] (9.5,-0.75) -- (10.2,-0.3);
\draw[-,dotted,dashed] (10.2,-0.3) -- (8.2,-0.3);
\draw[-,dotted,dashed] (8.2,-0.3) -- (7.5,-0.75);
%%
\draw[-,dotted] (7.5,0.75) -- (7.5,-0.75);
\draw[-,dotted] (9.5,0.75) -- (9.5,-0.75);
\draw[-,dotted] (10.2,1.2) -- (10.2,-0.3);
\draw[-,dotted,dashed] (8.2,1.2) -- (8.2,-0.3);
\end{tikzpicture}
\end{center}
\caption{\label{mult}Example of multiplication of two cubical types.}
\end{figure*}

This point is made precise in the following tentative denotation of types (to
be refined in the next section) which maps a type of dimension $n$ to an
$n$-dimensional cube. We represent such a cube syntactically as a binary tree
of maximum depth~$n$ with nodes of the form $\nodet{\cubt_1}{\cubt_2}$. In
such a node, $\cubt_1$ is the positive subspace and $\cubt_2$ (shaded in
gray) is the negative subspace along the first dimension. Each of these
subspaces is itself a cube of a lower dimension. The $0$-dimensional cubes
are plain sets representing the denotation of conventional first-order
types. We use $S$ to denote the denotations of these plain types. A
1-dimensional cube, $\nodet{S_1}{S_2}$, intuitively corresponds to the
difference $t_1 - t_2$ of the two types whose denotations are $S_1$ and $S_2$
respectively. The type can be visualized as a ``line'' with polarized
endpoints connecting the two points~$S_1$ and $S_2$. A full 2-dimensional
cube, $\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$, intuitively
corresponds to the iterated difference of the appropriate types
$(t_1-t_2)-(t_3-t_4)$ where the successive ``colors'' from the outermost box
encode the sign. The type can be visualized as a ``square'' with polarized
corners connecting the two lines corresponding to $(t_1-t_2)$ and
$(t_3-t_4)$. (See Fig.~\ref{mult} which is further explained after we discuss
multiplication below.)

@@

%%%%%%%%%%%%%%%%%%%%
\subsection{Type Isomorphisms: Paths to the Rescue}

Our proposed semantics of types identifies several structurally different
types such as $(1+(1+1))$ and $((1+1)+1)$. In some sense, this is innocent as
the types are isomorphic. However, in the operational semantics discussed in
the next section, we make the computational content of such type isomorphisms
explicit. Some other isomorphic types like $(t_1*t_2)$ and $(t_2*t_1)$ map to
different cubes and are \emph{not} identified: explicit isomorphisms are
needed to mediate between them. We therefore need to enrich our model of
types with isomorphisms connecting types we deem equivalent.

So far, our types are modeled as cubes which are really sets indexed by
polarities. An isomorphism between $(t_1*t_2)$ and $(t_2*t_1)$ requires
nothing more than a pair of set-theoretic functions between the spaces that
compose to the identity. What is much more interesting are the isomorphisms
involving the empty type~0. In particular, if negative types are to be
interpreted as their name suggests, we must have an isomorphism between
$(t-t)$ and the empty type 0. Semantically the former denotes the ``line''
$\nodet{\cubt}{\cubt}$ and the latter denotes the empty set. Their
denotations are different and there is no way, in the world of plain sets, to
express the fact that these two spaces should be identified. What is needed
is the ability to \emph{contract} the \emph{path} between the endpoints of
the line to the trivial path on the empty type. This is, of course, where the
ideas of homotopy (type) theory enter the development.

Consider the situation above in which we want to identify the spaces
corresponding to the types $(1-1)$ and the empty type:
\begin{center}
\begin{tikzpicture}
\node[above] at (0,0) {\pp};
\node[left] at (-0.2,0) {$1$};
\draw[fill] (0,0) circle [radius=0.05];
\node[above] at (2.6,0) {\mm};
\node[right] at (2.8,0) {$1$};
\draw[fill] (2.6,0) circle [radius=0.05];
\draw[->,red] (0.1,0) -- (2.5,0);
\draw[->,red] (-0.1,-0.1) arc (135:405:0.2);
\node[below,red] at (0,-0.4) {$\refl{1}$};
\draw[->,red] (2.5,-0.1) arc (135:405:0.2);
\node[below,red] at (2.6,-0.4) {$\refl{1}$};
\node[above] at (1.3,0) {$\seg{1}$};
\node[right] at (1.3,-0.5) {$q$};
%%
\node[below] at (1.3,-1.5) {$\emptyset$};
\draw[fill] (1.3,-1.5) circle [radius=0.05];
\draw[->,red] (1.4,-1.4) arc (-45:225:0.2);
\node[right,red] at (1.5,-1.3) {$\refl{0}$};
%% 
\draw[->,double,blue] (1.3,-0.1) -- (1.3,-1);
\end{tikzpicture}
\end{center}
The top of the figure is the 1-dimensional cube representing the type $(1-1)$
as before except that we now add a path $\seg{1}$ to connect the two
endpoints. (Note that previously, the dotted lines in the figures were a
visualization aid and were \emph{not} meant to represent paths.) We also make
explicit the trivial identity paths from every space to itself.  The bottom
of the figure is the 0-dimensional cube representing the empty type. The path
$\seg{1}$ identifies the two occurrences of 1. To express the
equivalence of $(1-1)$ and 0, we add a 2-path, i.e. a path between paths,
that asserts connects the path $\seg{1}$ to the trivial path
$\refl{0}$ on the empty type. That effectively makes the two points
``disappear.''  Surprisingly, that is everything that we need. The extension
to higher dimensions just ``works'' because paths in homotopy type theory
have a rich structure. We explain the details below.

%%%%%%%%%%%%%%%%%%%%
\subsection{The Universe of Types}

We describe the construction of our universe of types. Our starting point is
the construction in the previous section which we summarize again. We start
with all finite sets built from the empty set, a singleton set, disjoint
unions, and cartesian products. All these sets are thought of being indexed
by the empty sequence of polarities $\epsilon$. We then constructs new spaces
that consist of pairs of finite sets $\nodet{S_1}{S_2}$ indexed by positive
and negative polarities. These are the 1-dimensional spaces. We iterate this
construction to the limit to get $n$-dimensional cubes for all natural
numbers $n$.

At this point, we switch from viewing the universe of types as an
unstructured collection of spaces to a viewing it as a \emph{groupoid} with
explicit \emph{paths} connecting spaces that we want to consider as
equivalent. We itemize the paths we add next:
\begin{itemize}
\item The first step is to identify each space with itself by adding trivial
  identity paths $\refl{\cubt}$ for each space $\cubt$;
\item Then we add paths $\seg{\cubt}$ that connect all occurrences of the
  same type $\cubt$ in various positions in $n$-dimensional cubes. For
  example, the 1-dimensional space corresponding to $(1-1)$ would now include
  a path connecting the two endpoints at the end of the previous section;
\item We then add paths for witnessing the usual type isomorphisms between
  finite types such as associativity and commutativity of sums and
  products. The complete list of these isomorphisms is given in the next
  section.
\item Finally, we add \emph{2-paths} between $\refl{0}$ and any path
  $\seg{\cubt}$ whose endpoints are of opposite polarities, i.e., of
  polarities $s$ and $\negp{s}$ where:
  \[\begin{array}{rcl}
  \negp{\epsilon} &=& \epsilon \\
  \negp{+s} &=& -\negp{s} \\
  \negp{-s} &=& +\negp{s}
  \end{array}\]
\item The groupoid structure forces other paths to be added as described in
  detail in the recent book on Homotopy Type Theory~\cite{hottbook}. In
  particular, for every path $p$, the groupoid structure requires an inverse
  path $p^{-1}$ in the opposite direction; and for every two paths with
  common ending and starting point, a path corresponding to the composition
  of the two paths is included and so on. These paths obey various coherence
  conditions that ensure for example that composing a path and its inverse is
  equivalent to the trivial path, and that composition is associative, up to
  path equivalences at higher levels.
\end{itemize}

\begin{verbatim}
be careful about that 2-path. Are we 
allowed to do that given that the 
empty type is, well, empty.

identical to a single point. It should 
then follow that a square where one edge 
contracts to a point would itself contract 
to a line, and a 3-dimensional cube 
in which one face contracts to a point 
would itself contract to a square, etc.

So the story is that as we are building
the denotations of types, we add these
paths making up the cubes. The only
case where we add new paths is multiplication
(see Fig.1 again). We take the paths p in S1
and the paths q in S2 and build paths (p,q)
in the product space.

@@

Now that the type structure is defined, we turn our attention to the notion
of values. Intuitively, a value of the $n$-dimensional type $\tau$ is an
element of one of the sets located in one of the corners of the
$n$-dimensional cube denoted by $\tau$. Thus to specify a value, we must
first specify one of the corners of the cube (or equivalently one of the
leaves in the binary tree representation) which can easily be done using a
sequence $s$ of $+$ and $-$ polarities indicating how to navigate the cube in
each successive dimension starting from a fixed origin to reach the desired
corner. We write $v^{s}$ for the value $v$ located at corner $s$ of the cube
associated with its type. We use $\epsilon$ for the empty sequence of
polarities and identify $v$ with $v^\epsilon$. Note that the polarities
doesn't completely specify the type since different types like $(1+(1+1))$
and $((1+1)+1)$ are assigned the same denotation. What the path $s$ specifies
is the \emph{polarity} of the value, or its ``orientation'' in the space
denoted by its type. Formally:
\[\begin{array}{c}
\infer{() : 1}{} \\
\qquad
\infer[\textit{neg}]{v^{\negp{s}} : - \tau}{v^s : \tau} 
\\
\infer[\textit{left}]{(\inl{v})^{s} : \tau_1 + \tau_2}{v^{s} : \tau_1}
\qquad
\infer[\textit{right}]{(\inr{v})^{s} : \tau_1 + \tau_2}{v^{s} : \tau_2} 
\\
\infer[\textit{prod}]{(v_1,v_2)^{s_1 \cdot s_2} : \tau_1 * \tau_2}
      {v_1^{s_1} : \tau_1 & v_2^{s_2} : \tau_2} 
\end{array}\]
The rules \textit{left} and \textit{right} reflect the fact that sums do not
increase the dimension. Note that when $s$ is $\epsilon$, we get the
conventional values for the 0-dimensional sum type. The rule \textit{prod} is
the most involved one: it increases the dimension by \emph{concatenating} the
two dimensions of its arguments. For example, if we pair $v_1^{\pp}$ and
$v_2^{\mm\pp}$ we get $(v_1,v_2)^{\pp\mm\pp}$. (See Fig.~\ref{mult} for the
illustration.) Note again that if both components are 0-dimensional, the pair
remains 0-dimensional and we recover the usual rule for typing values of
product types. The rule \textit{neg} uses the function below which states
that the negation of a value $v$ is the same value $v$ located at the
``opposite'' corner of the cube.

@@

Connection to our work on univalence for finite types. We didn't have to rely
on sets for 0-dimensional types. We could have used groupoids again.
